Precise predictions for same-sign W-boson scattering at the LHC

### 2018-03-14 Draft/article_before_qsh.tex
ncluded

  MCnet-18-06, Nikhef/2018-012, UWTHPH-2018-12,

  VBSCan-PUB-01-18, ZU-TH-08/18}
\end{abstract}
\thispagestyle{empty}
\vfill
\newpage
\setcounter{page}{1}

@@

    \item The program {\sc Bonsay}~\cite{Dittmaier:2018zzz} consists of a general-purpose Monte Carlo integrator and matrix elements taken from different sources:
    Born matrix elements are adapted from the program {\sc Lusifer}~\cite{Dittmaier:2002ap}, which have been generalized to calculate also real matrix elements.
    Virtual matrix elements have been calculated using an in-house matrix element generator.
    One loop integrals are evaluated using the {\sc Collier} library~\cite{Denner:2014gla,Denner:2016kdg}.
    For the fiducial cross sections it uses the VBS approximation at LO and NLO.
    The virtual corrections are additionally approximated using a double-pole approximation.
    For more inclusive cross sections at LO the exact matrix elements ($s$-channels, interferences) can also be used.

@@

      \item {\sc VBFNLO}~\cite{Arnold:2008rz, Arnold:2011wj, Baglio:2014uba} is a flexible
        parton-level Monte Carlo for processes with EW bosons. It
        allows the calculation of VBS processes at NLO QCD in the VBS
        approximation, with process IDs between 200 and 290. Same-sign
        $\PW^+\PW^+jj$ production is provided via the process ID 250. The corresponding
        $s$-channel contributions are available separately as tri-boson processes with
        semi-leptonic decays, with process IDs between 401 and 492. For the
        final state studied in this article, only $\PW^+\PW^+\PW^-$
        production with a hadronically decaying $\PW^-$ boson, process ID 432,
        can contribute. These can simply be added on top of the VBS
        contribution. Interferences between the two are therefore neglected.
        The usage of leptonic tensors in the calculation, pioneered in
        Ref.~\cite{Jager:2006zc}, thereby leads to a significant speed improvement over
        automatically generated code.  Besides the SM, also a variety of
        new-physics models including anomalous couplings of the Higgs and gauge
        bosons can be simulated.

@@

We now present the results of predictions matched to parton showers.
The total rates within VBS cuts are displayed in Tables~\ref{tab:PSratesLO} and
\ref{tab:PSratesNLO}, at LO and NLO
accuracy respectively. For {\sc MG5\_aMC},
the numbers with $\Gamma_{\rm resc}$ are rescaled to
take into account the width effects described in the above paragraph. At NLO accuracy, for {\sc MG5\_aMC} + {\sc Pythia8} and {\sc VBFNLO}-{\sc Dipole}, we also quote
theoretical uncertainties.
For the former, we show both PDF and scale uncertainties,\footnote{A preliminary study on PDF uncertainties in VBS has appeared
in Ref.~\cite{Schwan:2017yy}.} obtained via exact reweighting~\cite{Frederix:2011ss} by varying independently the renormalisation and factorisation
scales by a factor of two around the central value, Eq.~(\ref{eq:defscale}) (nine-point variations).
For the latter, we show the
three-point scale uncertainties, obtained by considering correlated variations of the renormalisation and factorisation scales. Theory uncertainties should have very little dependence on the tool employed.
We observe that, once the width effect is taken into
account, total rates from different tools agree within some per cents, both at LO and NLO. Larger discrepancies, however, will appear for differential observables, which we are going to discuss in
the following. Theory uncertainties on the total rates are very small, regardless of whether scale variations are estimated with 
independent or correlated variations of the renormalisation and factorisation scales.
Concerning differential distributions, for each observable we will display results in two plots, shown side-by-side. In the plot on the left (right), (N)LO+PS predictions are shown
with different colours in the main frame. In the inset, these predictions are compared in both cases with a fixed-order prediction at NLO accuracy (obtained with
{\sc VBFNLO}, \emph{i.e.}\ the VBS approximation with $s$-channel).
For the differential observables, the {\sc MG5\_aMC} predictions are \emph{not} rescaled to compensate for the width effect mentioned above. As for the table, we show theoretical uncertainties for the NLO+PS samples
obtained with {\sc VBFNLO} and {\sc MG5\_aMC}:
again, for the first the band corresponds to three-point variations, while for the second the darker (lighter) band corresponds to nine-point
scale variations (plus PDF uncertainties, linearly added).

The first observable we investigate is the exclusive jet multiplicity, shown in Fig.~\ref{fig:PSnjet}. Looking at the LO+PS predictions, one can appreciate that the
main effects are driven by the parton shower that is employed ({\sc Herwig++/7} or {\sc Pythia8}), with the clear tendency of producing more radiation for the latter,
leading to higher jet multiplicities. Differences among tools that employ the same parton shower are typically smaller, and can be traced back to different values of the
initial scale of the parton shower (the {\tt scalup} entry of the Les Houches Event (LHE) file \cite{Alwall:2006yp,Butterworth:2010ym}). This event-by-event number corresponds
to the maximum hardness (translated into the shower evolution variable) of the radiation that
can be generated by the shower. \footnote{At LO, the choice of such a scale is arbitrary and usually driven by common sense,
as it is the case for the factorisation and renormalisation scales.
    At NLO, one has the freedom to change the shower scale without losing formal NLO accuracy within the {\sc MC@NLO} matching,
    provided the Monte Carlo counterterms are also consistently
updated. In the {\sc Powheg} matching, the shower scale of the so-called $\tilde B$ events is fixed to the transverse momentum of the radiation
generated according the {\sc Powheg} Sudakov factor, while it can be changed in the remnants events.}
The main effect of NLO corrections for this (rather inclusive) observable is to stabilise the predictions for the two-jet bin, where discrepancies
among tools are reduced to about $10\%$. For the three-jet bin, which is described only at LO accuracy, differences among tools remain large: the largest rate is predicted by
{\sc MG5\_aMC}, while the smallest rate is predicted by the {\sc Powheg-Box}, both matched to {\sc Pythia8}. Despite the fact that the same parton shower is employed, the way emissions are treated
is different among the two tools. In particular, for the {\sc Powheg-Box}, the first emission is generated with an internal Sudakov form factor (the
prediction dubbed {\sc Powheg-LHE} corresponds to stopping after the first emission), while for {\sc MG5\_aMC} there is an
interplay between the real-emission matrix element and the shower emission. For this observable we also show the prediction obtained
with {\sc MG5\_aMC+Pythia8} by reducing the shower-starting scale by a factor 2 with respect to the default value, dubbed 
{\sc MG5\_aMC+Py8, $Q_{sh}/2$}.\footnote{The reduction of the shower scale for {\sc MG5\_aMC+Herwig++} gives no visible effect on any of the observables 
discussed in this work.} The main effect of reducing the shower scale is that events migrate from the three-jet bin into the two-jet bin, i.e. less radiation is generated. The
amount of this effect on the jet rate is +4\% (-8\%) on the two (three) jet bin, while the total rate within cuts is left unchanged.

@@

The next observable that we study is the invariant mass of the two tagging jets, shown in Fig.~\ref{fig:PSmjj}. For this observable, both at LO+PS and NLO+PS,
the spread of predictions matched with parton shower is rather small
($\lesssim 10\%$, if one compensates for the $6\%$ width effect for {\sc MG5\_aMC}).
The LO+PS predictions tend to be significantly softer than the fixed NLO one, with an effect of
about -30\% at the end of the displayed range. At NLO+PS, this effect is mitigated, owing to the better description of the first QCD emission which is now driven by the real-emission matrix element. For this observable (and all the others which are NLO accurate) the effect of reducing the shower-scale is 
negligible, hence the corresponding histogram will not be shown.

@@

Finally, focusing on the third jet, we conclude the list of differential observables by showing the Zeppenfeld variable defined in Eq.~\eqref{eq:Zeppenfeld}, Fig.~\ref{fig:PSz3}. This
variable is closely related to the third jet rapidity, and small (large) values of $z$ correspond to central (peripheral) rapidities. In general, for observables which involve the third jet, one
can clearly see a degradation of the agreement among the various tools, because of the poorer perturbative description of these observables. The Zeppenfeld variable is
a striking example: both at LO and NLO, the tendency of {\sc Pythia8} to generate more hard and central radiation, corresponding to low values of $z$,
is clearly visible. Such an effect, which is related to the way {\sc Pythia8} deals with the recoil of the radiation in VBF(VBS)-type processes,
can be mitigated by setting {\tt SpaceShower:dipole\-Recoil = on} in the {\sc Pythia8} input file.\footnote{This requires version $\ge8.230$.
Note that such a setting is not compatible with the NLO matching
in {\sc MG5\_aMC} (but it is compatible with the {\sc Powheg} matching). Also, this setting has other
effects, though smaller, on the rapidity spectra of the two hardest jets.} It is interesting to notice that
the effect survives beyond the first emission, as it can be observed by comparing {\sc Powheg-LHE} with {\sc Powheg+ Pythia8}, and that
it is only marginally attenuated when the shower scale is reduced. A similar
behaviour of {\sc Pythia8}
has also been observed in the study of EW production of a $\PZ$ boson in association with two jets (see the recent CMS measurement,
Ref.~\cite{Sirunyan:2017jej} Figure 12), where the experimental data seem to prefer the description by {\sc Herwig++}.
The central enhancement
is a bit mitigated if NLO+PS tools are used (compare LO+PS and NLO+PS from {\sc MG5\_aMC+Pythia8} with the fixed-NLO prediction), however even at NLO+PS the central region
($z_{j_3}<0.5$) is cursed by huge differences between tools. Large differences, reaching a factor 2, persist also away from the central region. These findings are consistent with behaviour displayed in Refs.~\cite{Jager:2011ms,Jager:2012xk,Jager:2013mu,Jager:2013iza,Schissler:2013nga} where the behaviour of NLO matching in VBS processes has been reported.\\

In conclusion, the comparison of tools including matching to parton shower clearly shows the benefits of the inclusion of NLO corrections: for most observables described
effectively at NLO accuracy differences between tools are at (or below) the $10\%$ level.
Some exceptions exist, \emph{e.g.}\ the rapidity separation of the two tagging jets, which
on the one hand clearly suggest not to rely on a single tool/parton shower, and on the other make it worth to investigate more in detail the way QCD radiation is
generated, \emph{e.g.}\ when fully-differential computations at NNLO will become available (for VBF Higgs production, see Refs.~\cite{Cacciari:2015jma, Cruz-Martinez:2018rod}). It is a remarkable fact that, even for those observables that display small discrepancies,
the theoretical uncertainty obtained via scale variations (renormalization, factorization and shower scale) 
systematically underestimates the spread of predictions. We note 
that in the only VBF process where NNLO corrections are known, i.e. VBF Higgs production~\cite{Cacciari:2015jma, Cruz-Martinez:2018rod}, 
the NLO scale-uncertainty band does not include the NNLO prediction. This suggest that the NLO scale variation 
underestimates the size of the perturbative uncertainty. Again, this stresses the need
 to employ at least two different tools in order to obtain a conservative estimate of theoretical uncertainties. Finally, the size of discrepancies for observables that are described at a lower perturbative accuracy, notably those related to the third jet, suggest that
experimental analyses should rely as little as possible on those observables and, in any case, use conservative estimate of the theory
uncertainties. On the one hand, in order to improve the
description of these observables, a simulation of VBS+j at NLO accuracy, currently unavailable but within the reach of modern
automated tools, is certainly desirable.
On the other hand, measurements of processes with similar
color flow (EW production of a single vector boson plus jets,
VBF, \ldots) can certainly help in order to discriminate which tools perform better in the comparison with data~\cite{Aaboud:2017emo,Sirunyan:2017jej}.

>>>>

  Nikhef/2018-012, UWTHPH-2018-12, VBSCan-PUB-01-18,

  ZU-TH-08/18}
\end{abstract}
\thispagestyle{empty}
\vfill
\newpage
\setcounter{page}{1}

@@

    \item The program {\sc Bonsay} consists of a general-purpose Monte Carlo integrator written by Christopher Schwan and matrix elements taken from different sources:
    Born matrix elements are adapted from the program {\sc Lusifer}~\cite{Dittmaier:2002ap}, real matrix elements are written by Marina Billoni, and virtual matrix elements by Stefan Dittmaier.
    One loop integrals are evaluated using the {\sc Collier} library~\cite{Denner:2014gla,Denner:2016kdg}.
    For the fiducial cross sections it uses the VBS approximation at LO and NLO.
    The virtual corrections are additionally approximated using a double-pole approximation.
    For more inclusive cross sections at LO the exact matrix elements ($s$-channels, interferences) can also be used.

@@

      \item {\sc VBFNLO}~\cite{Arnold:2008rz, Arnold:2011wj, Baglio:2014uba} is a flexible
        parton-level Monte Carlo for processes with EW bosons. It
        allows the calculation of VBS processes at NLO QCD in the VBS
        approximation, with process IDs between 200 and 290. The corresponding
        $s$-channel contributions are available separately as tri-boson processes with
        semi-leptonic decays, with process IDs in the 400 range\ZM{which range is it?}. These can simply
        be added on top of the VBS contribution. Interferences between the two are therefore neglected.
        The usage of leptonic tensors in the calculation, pioneered in
        Ref.~\cite{Jager:2006zc}, thereby leads to a significant speed improvement over
        automatically generated code.  Besides the SM, also a variety of
        new-physics models including anomalous couplings of the Higgs and gauge
        bosons can be simulated.

@@

We now present the results of predictions matched to parton showers.
The total rates within VBS cuts are displayed in Tables~\ref{tab:PSratesLO} and
\ref{tab:PSratesNLO}, at LO and NLO
accuracy respectively. For {\sc MG5\_aMC},
the numbers with $\Gamma_{\rm resc}$ are rescaled to
take into account the width effects described in the above paragraph. At NLO accuracy, for {\sc MG5\_aMC} + {\sc Pythia8} and {\sc VBFNLO}-{\sc Dipole}, we also quote
theoretical uncertainties.
For the former, we show both PDF and scale uncertainties,\footnote{A preliminary study on PDF uncertainties in VBS has appeared
in Ref.~\cite{Schwan:2017yy}.} obtained via exact reweighting~\cite{Frederix:2011ss} by varying independently the renormalisation and factorisation
scales by a factor of two around the central value, Eq.~(\ref{eq:defscale}) (nine-point variations).
For the latter, we show the
three-point scale uncertainties, obtained by considering correlated variations of the renormalisation and factorisation scales. Theory uncertainties should have very little dependence on the tool employed.
We observe that, once the width effect is taken into
account, total rates from different tools agree within some or few per cents, respectively at LO and NLO. Larger discrepancies, however, will appear for differential observables, which we are going to discuss in
the following. Theory uncertainties on the total rates are very small, regardless of whether scale variations are estimated with 
independent or correlated variations of the renormalisation and factorisation scales.
Concerning differential distributions, for each observable we will display results in two plots, shown side-by-side. In the plot on the left (right), (N)LO+PS predictions are shown
with different colours in the main frame. In the inset, these predictions are compared in both cases with a fixed-order prediction at NLO accuracy (obtained with
{\sc VBFNLO}, \emph{i.e.}\ the VBS approximation with $s$-channel).
For the differential observables, the {\sc MG5\_aMC} predictions are \emph{not} rescaled to compensate for the width effect mentioned above. As for the table, we show theoretical uncertainties for the NLO+PS samples
obtained with {\sc VBFNLO} and {\sc MG5\_aMC}:
again, for the first the band corresponds to three-point variations, while for the second the darker (lighter) band corresponds to nine-point
scale variations (plus PDF uncertainties, linearly added).

The first observable we investigate is the exclusive jet multiplicity, shown in Fig.~\ref{fig:PSnjet}. Looking at the LO+PS predictions, one can appreciate that the
main effects are driven by the parton shower that is employed ({\sc Herwig++/7} or {\sc Pythia8}), with the clear tendency of producing more radiation for the latter,
leading to higher jet multiplicities. Differences among tools that employ the same parton shower are typically smaller, and can be traced back to different values of the
initial scale of the parton shower (the {\tt scalup} entry of the Les Houches Event (LHE) file \cite{Alwall:2006yp,Butterworth:2010ym}). This event-by-event number corresponds
to the maximum hardness (translated into the shower evolution variable) of the radiation that
can be generated by the shower. \footnote{At LO, the choice of such a scale is arbitrary and usually driven by common sense,
as it is the case for the factorisation and renormalisation scales.
    At NLO, one has the freedom to change the shower scale without losing formal NLO accuracy within the {\sc MC@NLO} matching,
    provided the Monte Carlo counterterms are also consistently
updated. In the {\sc Powheg} matching, the shower scale of the so-called $\tilde B$ events is fixed to the transverse momentum of the radiation
generated according the {\sc Powheg} Sudakov factor, while it can be changed in the remnants events.}
The main effect of NLO corrections for this (rather inclusive) observable is to stabilise the predictions for the two-jet bin, where discrepancies
among tools are reduced to about $10\%$. For the three-jet bin, which is described only at LO accuracy, differences among tools remain large: the largest rate is predicted by
{\sc MG5\_aMC}, while the smallest rate is predicted by the {\sc Powheg-Box}, both matched to {\sc Pythia8}. Despite the fact that the same parton shower is employed, the way emissions are treated
is different among the two tools. In particular, for the {\sc Powheg-Box}, the first emission is generated with an internal Sudakov form factor (the
prediction dubbed {\sc Powheg-LHE} corresponds to stopping after the first emission), while for {\sc MG5\_aMC} there is an
interplay between the real-emission matrix element and the shower emission.

@@

The next observable that we study is the invariant mass of the two tagging jets, shown in Fig.~\ref{fig:PSmjj}. For this observable, both at LO+PS and NLO+PS,
the spread of predictions matched with parton shower is rather small
($\lesssim 10\%$, if one compensates for the $6\%$ width effect for {\sc MG5\_aMC}).
The LO+PS predictions tend to be significantly softer than the fixed NLO one, with an effect of
about -30\% at the end of the displayed range. At NLO+PS, this effect is mitigated, owing to the better description of the first QCD emission which is now driven by the real-emission matrix element.

@@

Finally, focusing on the third jet, we conclude the list of differential observables by showing the Zeppenfeld variable defined in Eq.~\eqref{eq:Zeppenfeld}, Fig.~\ref{fig:PSz3}. This
variable is closely related to the third jet rapidity, and small (large) values of $z$ correspond to central (peripheral) rapidities. In general, for observables which involve the third jet, one
can clearly see a degradation of the agreement among the various tools, because of the poorer perturbative description of these observables. The Zeppenfeld variable is
a striking example: both at LO and NLO, the tendency of {\sc Pythia8} to generate more hard and central radiation, corresponding to low values of $z$,
is clearly visible. Such an effect, which is related to the way {\sc Pythia8} deals with the recoil of the radiation in VBF(VBS)-type processes,
can be mitigated by setting {\tt SpaceShower:dipole\-Recoil = on} in the {\sc Pythia8} input file.\footnote{This requires version $\ge8.230$.
Note that such a setting is not compatible with the NLO matching
in {\sc MG5\_aMC} (but it is compatible with the {\sc Powheg} matching). Also, this setting has other
effects, though smaller, on the rapidity spectra of the two hardest jets.} It is interesting to notice that
the effect survives beyond the first emission, as it can be observed by comparing {\sc Powheg-LHE} with {\sc Powheg+ Pythia8}. A similar
behaviour of {\sc Pythia8}
has also been observed in the study of EW production of a $\PZ$ boson in association with two jets (see the recent CMS measurement,
Ref.~\cite{Sirunyan:2017jej} Figure 12), where the experimental data seem to prefer the description by {\sc Herwig++}.
The central enhancement
is a bit mitigated if NLO+PS tools are used (compare LO+PS and NLO+PS from {\sc MG5\_aMC+Pythia8} with the fixed-NLO prediction), however even at NLO+PS the central region
($z_{j_3}<0.5$) is cursed by huge differences between tools. Large differences, reaching a factor 2, persist also away from the central region. These findings are consistent with behaviour displayed in Refs.~\cite{Jager:2011ms,Jager:2012xk,Jager:2013mu,Jager:2013iza,Schissler:2013nga} where the behaviour of NLO matching in VBS processes has been reported.\\

In conclusion, the comparison of tools including matching to parton shower clearly shows the benefits of the inclusion of NLO corrections: for most observables described
effectively at NLO accuracy differences between tools are at (or below) the $10\%$ level.
Some exceptions exist, \emph{e.g.}\ the rapidity separation of the two tagging jets, which
on the one hand clearly suggest not to rely on a single tool/parton shower, and on the other make it worth to investigate more in detail the way QCD radiation is
generated, \emph{e.g.}\ when fully-differential computations at NNLO will become available (for VBF Higgs production, see Refs.~\cite{Cacciari:2015jma, Cruz-Martinez:2018rod}). It is a remarkable fact that, even for those observables that display small discrepancies,
the theoretical uncertainty obtained via scale variations systematically underestimates the spread of predictions. We note 
that in the only VBF process where NNLO corrections are known, i.e. VBF Higgs production~\cite{Cacciari:2015jma, Cruz-Martinez:2018rod}, 
the NLO scale-uncertainty band does not include the NNLO prediction. This suggest that the NLO scale variation 
underestimates the size of the perturbative uncertainty. Again, this stresses the need
 to employ at least two different tools in order to obtain a conservative estimate of theoretical uncertainties. Finally, the size of discrepancies for observables that are described at a lower perturbative accuracy, notably those related to the third jet, suggest that
experimental analyses should rely as little as possible on those observables and, in any case, use conservative estimate of the theory
uncertainties. On the one hand, in order to improve the
description of these observables, a simulation of VBS+j at NLO accuracy, currently unavailable but within the reach of modern
automated tools, is certainly desirable.
On the other hand, measurements of processes with similar
color flow (EW production of a single vector boson plus jets,
VBF, \ldots) can certainly help in order to discriminate which tools perform better in the comparison with data~\cite{Aaboud:2017emo,Sirunyan:2017jej}.
